Quantifying and controlling biases in dark matter halo concentration estimates

### 2016-08-25 paper/ms.tex
muchos edits, casi listo

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
In this paradigm the sizes and dynamics of galaxies are driven by
the DM distribution inside a halo. This has motivated the detailed study
a DM halo internal structure in the last three decades.

The internal DM distribution in a halo is usually parameterized
through the density profile.  In a first approximation this profile is
spherically symmetric; the density only depends on the radial
coordinate.  One of the most popular radial parameterizations is
the Navarro-Frenk-White (NFW) profile \citep{NFW}.  
This profile can be considered as universal \citep{Navarro2010}, assuming that 
one is not interested in the very central region of the dark matter halo where
galaxy formation takes place, and where the effects of baryon physics
on dark matter distribution are still unknown.
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  The ratio between
the scale radius and the halo virial radius $R_v$ is known as the
concentration $c=R_v/r_s$.

The concentration of the NFW profile provides a conceptual framework
to study simulated dark matter halos as a function of redshift and
cosmological parameters.  In many numerical studies
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014}
the results are summarized through the mass-concentration
relationship; that is the distribution of concentration values at a
fixed halo mass and redshift.  The success of such numerical
experiments rests on a reliable algorithm to estimate the
concentration.  Such an algorithm should provide unbiased results and
must be robust when applied to simulations of different numerical
resolution.

There are two established algorithms to estimate the concentration
parameter of a N-body dark matter halo.  The first method takes the
halo particles and bins them into logarithmic radii to estimate the
density in each bin, then it proceeds to fit the density as a function
of the radius to the NFW profile.  A second method uses an analytic
property of the NFW profile that relates the maximum of the ratio of
the circular velocity to the virial velocity, $V_{\rm circ}$/$V_{\rm
  vir}$.  The concentration can be then found as the root of an
algebraic equation dependent on this maximum value.

The first method is straightforward to apply but presents two
disadvantages.  First, it requires a large number of particles in
order to have a proper density estimate in each bin.  This makes the
method robust only for halos with at least $10^2$ particles.  The
second problem is that there is not a way to estimate the optimal
radial bin size, different choices may produce different results for
the concentration.

The second method solves the two problems mentioned above.  
It works with low particle numbers and does not involve data binning.  
However,
it effectively takes into account only a single data point and
discards the behavior of the ratio $V_{\rm circ}/V_{\rm vir}$ below
and above its maxima.  
Small fluctuations on the value of this maximum
can yield large perturbations on the estimated concentration
parameter.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{\emph{Left panel}. Average value of the fractional offset ($f_{\rm
    off}$ (Eq. \ref{eq:f_off}) between the concentration in a halo at 
  high resolution and its bootstrapped versions. 
  Thick (thin) lines correspond to massive halos in the Bolshoi (Via
  Lactea) simulation.
  For each case close to $10000$ bootstrapping samples are used.
  The density method noticeably overestimates the concentration up to
  a factor of $20\%$, 
  while the new method only underestimates the concentrations by less
  than $3\%$.
  \emph{Right panel.}
  1$\sigma$ uncertainties on the offset and the concentration
  estimates from the integrated mass method.  
  Lines show the width between the 14 and 86 percentiles
  of the $f_{\rm off}$ distribution at fixed particle number. 
  The lines include the results for the three methods: maximum
  velocity, density and integrated mass.
  Grey symbols correspond to the $1\sigma$ uncertainty on each of
  the halos in the Bolshoi Simulation as a result of the MCMC
  sampling. 
  To allow a fair comparison with $f_{\rm off}$, this uncertainty has
  been normalized by the preferred concentration value for each halo. 
  The $1\sigma$ bootstrapping uncertainty follow a $1/\sqrt{N_s}$
  behaviour. 
  \label{fig:downsampling}}
\end{figure*}

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.  The first method
tries to directly estimate the density profile.  It takes all the
particles in the halo and bins them in the logarithm of the radial
coordinate from the halo center.  Then, it estimates the density in
each logarithmic bin counting the particles and dividing by the
corresponding shell volume.  At this point is possible to make a
direct fit to the density as a function of the radial coordinate.
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).
This method has been most recently used by \cite{Klypin2016} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

@@

Here we propose a new method to estimate the concentration. 
It uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).  We build it from N-body data following the
next steps.  First, we define the center of the halo to be at the
position of the particle with the lowest gravitational potential.
Then we rank the particles by their increasing radial distance from
the center.  From this ranked list of $i=1,N$ particles, the total
mass at a radius $r_i$ is $M_i=i\times m_p$, where $r_i$ is the
position of the $i$-th particle and $m_p$ is the mass of a single
computational particle.

We divide the enclosed mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter.

@@

If the integrated mass values at different radii were independent from
each other we could write a likelihood distribution as ${\cal
  L}(c)\propto \exp(-\chi^2(c)/2)$ with

\begin{equation}
\chi^2(c)= \sum_{i=2}^{N-1}\frac{[\log m_i - \log m(< x_i;c)]^2}{\sigma_i^2},
\end{equation}
%
where $\sigma_i^2=\sigma_x^2(x_i)$, $m(<x_i;c)$ corresponds to the
values in Eq.(\ref{eq:profile}) at $x=x_i$ for a given value of the
concentration parameter $c$, and the $i$ index sums over all the
particles in the numerical profile. 
In this computation the particles $i=1$ and $i=N$ are discarded to
avoid divergent terms in the sum.

However, tests on the bootstrapping data show that using
$\sigma_i^2=\sigma_x^2(x_i)$, instead of the full inverse covariance
matrix, grossly overestimates $\chi^2(c)$,
providing small uncertainties around the best concentration value. 
To avoid the expensive computation and inversion of a full covariance
matrix we use the bootsrapping data to calibrate an effective
$\sigma_i^2\approx \sigma_{\rm eff}^2(x_i)$.

We impose two conditions on the approximate $\sigma_{\rm eff}^{2}$. It
must keep the dependence on $x$ that we have discovered for the
diagonal elements and must give similar curves of $\chi^{2}(c)$-$c$
around the minimum as the full covariance matrix.
We found that this effective $\sigma_{\rm eff}^2$ can be approximated
as

@@

We then use an Affine Invariant Markov Chain Monte Carlo (MCMC)
implemented in the python module {\texttt{emcee}} \citep{emcee} to sample the
likelihood function distribution.
From the $\chi^2$ distribution we find the optimal value of the
concentration and its associated uncertainty. Different choices for
$\sigma^2_{\rm eff}$ do not affect the optimal concentration value,
only its uncertainty.

@@

We use two very different simulations to test our methods.
The first data sample comes from the Bolshoi run, a cosmological
simulation that follows the non-linear evolution of a dark matter density field
sampled with $2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.  
The cosmological parameters use a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The data is publicly available at \url{http://www.cosmosim.org/}.  
Details about the structure of the database and the simulation can be
found in \citep{2013AN....334..691R}.

We use a halo sample containing all the halos located in a cubic
sub-volume of $100$ \hMpc\ on a side containing a total of $64531$
halos. 
From this sample we select all the halos at $z=0$ detected with a
Friends-of-Friends (FoF) algorithm with more than 300 particles,
meaning that the masses are in the interval $4\times 10^{10}\leq
M_{\rm FoF}/\hMsun \leq 10^{14}$.  The FoF algorithm used a linking
length of $0.17$ times the mean inter-particle distance.  
This choice translates into an overdensity $\Delta_h\sim 400-700$
dependent on the halo concentration \citep{More2011}.

From this set of particles we follow the procedure spelled out in
Section \ref{sec:method} with $\Delta_h=740$ (corresponding to $200$
times the critical density) to select an spherical region that we
redefine to be our halo.  
This choice makes that the overdensities are fully included inside the
original FoF particle group.   
On the interest of providing a fair comparison against the density
method we only report results from overdensities with at least $200$
particles ($2.6\times 10^{10}$\hMsun).

We also use public data from the Via Lactea simulation project
\citep{2008Natur.454..735D}.  
This simulation was run for a single isolated halo with a virial mass
of the order of $10^{12}$\hMsun using the parallel tree code PKDGRAV
\citep{2001PhDT........21S}.  
The simulation used $\sim 2\times 10^{8}$ particles to resolve this
region.
The cosmological parameters are different from those in the Bolshoi
simulation, with a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.238$ and a normalization of the power spectrum
$\sigma_8=0.74$. 
The data available to the public corresponds to a downsampled set 
of $10^5$ particles which corresponds to a particle mass of
$2.24\times 10^{7}$\hMsun.

@@

In our bootstrapping experiments we take halos sampled with at least
$10^{5}$ particles and subsample each halo by factors of $2$ up to
$10^{3}$ to measure the concentration every time the halo is
resampled. 
We use a two-sample Kolmogorov-Smirnov (KS) test to check that the
bootstrapped sets are a fair subsample of the parent halo.
Using the KS test we compare the list of radial distances from each
subsample against that of its parent halo.
We find that the resulting p-value distribution is flat.
This confirms that the radial distribution of particles in the
bootraped halo is consistent with coming from the distribution given
by the parent halo.

For every subsample we keep fixed the virial radius and the center
found for the high resolution halo.
Leaving the virial radius and center free in each bootstrapping
iteration has an effect smaller than $1\%$ in the concentration. 
We extract in total 10000 bootstrapping samples for the Bolshoi
and the Via Lactea simulations.
In the Bolshoi simulation we select 14 massive halos and create 700
subsamples for each one.  
For the Via Lactea simulation the same halo is subsampled 10000
times.

@@

Figure \ref{fig:downsampling} summarizes our results.  
The plot on the left shows the average value of $f_{\rm off}$ as a
function of particle number.  
As expected, for a large enough particle number, $N_{s}>4\times 10^3$,
the results of the three algorithms coincide below the $1\%$ level.
For a lower number of particles the results start to deviate.
At $200$ particles the velocity method overestimates the concentration
by a factor $14\%$ while the density method overestimates it by $20\%$.
Around the same sampling scale, the new algorithm shows a more stable
behaviour underestimating the concentration only by a factor of
$1\%$-$3\%$

The thin continous lines on the same panel shows a fit to the data
with the following functional form 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}}.
\label{eq:f_off_fit}
\end{equation} 
for the dataset shown in the panel we find $A=2842\pm 1900$,
$B=7.96\pm 0.54$ for the density method; $A=239\pm 131$,
$B=6.23\pm 0.43$ for the velocity method; $A=-0.46\pm 3.49$,
$B=0.79\pm 1.31$ for the integrated mass method.
The uncertainties are estimated from the covariance  matrix resulting
from the non-linear optimization procedure. The lines on Figure
\ref{fig:downsampling} are plotted using these best values.

The right panel in Figure \ref{fig:downsampling} shows different uncertainty
results.  
The lines show the difference between the 14 and 86 percentiles in the
$f_{\rm off}$ distribution at fixed mass. 
Each line corresponds to the three different methods to estimate the
concentration and the two simulations.
This shows that the bootstrapping technique can help us to assign a
$1\sigma$ uncertainty to the concentration values at a fixed $N_s$as 
%
\begin{equation}
\sigma_{c} = \frac{0.40}{\sqrt{N_s/200}}.
\end{equation}
%
An interesting consequence of this result is that $\sigma_c=1$ for
$N_s=32$ particles, independent of the method used to estimate the
concentration.

The circles in the same Figure show the $1\sigma$ uncertainty as
estimated from the MCMC chains using the integrated mass method on all
the relaxed halos in the Bolshoi simulation sample. 
To allow for a fair comparison with the bootstrapping results, this
uncertainty was normalized by the concentration value.
Below the range $N_{s}<10^5$ the MCMC uncertainty is lower than the
bootstrapping estimate, with a weak dependence on the particle
number.

@@

We now compare the results for the mass-concentration
relationship obtained with the three different algorithms.
This helps us to study the concentration results
for the new algorithm and, more importantly, showcase a probable
consequence of the biases detected through bootstrapping.

Figure \ref{fig:concentration} shows the mass-concentration
relationship for the density, velocity and integrated mass method.
The left panel shows the results as they are produced by each of the
algorithms. The thin dashed line marks the trend reported by
\citep{Prada2012} using the velocity method.

We notice first that the results from the new algorithm follow very
closely the velocity algorithm at high masses ($M_h>10^{12}\hMsun$ or
equivalently for $>4\times10^3$ particles). 
For lower masses there is a difference between the median of the two
methods, but they are still consistent within the statistical
uncertainties.

@@

@@

In this letter we used bootstrapping to quantify the biases on
concentration estimates. 
We found that methods commonly used in the literature present a large
bias by overestimating concentrations by factors of $15\%$-$20\%$ for
halos with $200$ particles, or $7\%$-$10\%$ for halos with $500$
particles.  
This procedure provides a robust technique to quantify the bias in
concentration estimates with the advantage that it works without
having to run new simulations.

These results motivated us to introduce a method with a robust
performance at low particle numbers.
This method is based on the integrated mass profile. 
The results of the algorithm on the same data sample showed a bias  of
$< 3\%$ for halos with $200$ particles and less than $1\%$ for halos
with $500$ particles or more.  
To keep the bias of the velocity and density methods below $2\%$ only halos
with at least $\sim 4000$ particles should be taken into account.

Measuring the mass concentration relationship the three methods are in
broad agreement within the statistical  uncertainties, although there
are some noticeable differences. 
For instance, the density method produces systematically higher
concentrations by a factor of $15\%$ compared to the velocity method.
This systematic offset has been reported before with the same dataset 
\citep{Prada2012} and with different simulations \citep{Klypin2016}
without any conclusive explanation for its origin. 
A second difference is that the velocity and integrated mass methods
start to differ for masses below $10^{12}\hMsun$ ($\sim 4000$
particles).   
We found that correcting the mean concentration by the mean bias
factor found through bootstrapping brings these two techniques into
agreement.

These results show that using the integrated mass profile to find the
Dark Matter halo concentration is a tool deserving deeper scrutiny.   
Further tests with larger simulated volumes, varying numerical
resolution and different density profiles are the next natural step to
explore the full potential of this new method.

@@

 The authors acknowledge the technical support from the new
 high-performance computing facility at Uniandes. JEFR acknowledges
 financial support from Vicerrector\'ia de Investigaciones at Uniandes
 through a FAPA project. JCMC acknowledges financial support from
 ``Estrategia de  sostenibilidad 2014-2015, Universidad de
 Antioquia''. The authors also thank Tom\'as Verdugo, Stefan
 Gottloeber and Nelson Padilla for their comments on earlier version
 of this work.

\bibliographystyle{apj}
%\bibliography{references}
\begin{thebibliography}{15}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

@@

@@

>>>>

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
In this paradigm the sizes and dynamics of galaxies are driven by
the halo internal DM distribution.

The internal DM distribution in a halo is usually parameterized
through the density profile.  
In a first approximation this profile is spherically symmetric; the
density only depends on the radial coordinate.  
One of the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.   
This profile can be considered as universal \citep{Navarro2010}, assuming that 
one is not interested in the very central region where galaxy
formation takes place, and where the effects of baryon physics on the
dark matter distribution are still unknown. 
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The concentration of the NFW profile provides a conceptual framework
to study simulated dark matter halos as a function of redshift and
cosmological parameters.  
Numerical studies
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014,Ludlow2016,Klypin2016}
summarized their results through the mass-concentration
relationship; that is, the distribution of concentration values at a
fixed halo mass and redshift.  
The success of such numerical experiments rests on a reliable
algorithm to estimate the concentration.  
Such an algorithm should provide unbiased results and must be robust
when applied at varying numerical resolution.

There are two established algorithms to estimate the concentration
parameter.
The first method takes the halo particles and bins them into
logarithmic radii to estimate the density in each bin, then it
proceeds to fit the density as a function of the radius.  
A second method uses an analytic property of the NFW profile that
relates the maximum of the ratio of the circular velocity to the
virial velocity, $V_{\rm circ}$/$V_{\rm   vir}$.  The concentration
can be then found as the root of an algebraic equation dependent on
this maximum value.

The first method is straightforward to apply but presents two
disadvantages.  First, it requires a large number of particles in
order to have a proper density estimate in each bin.  This makes the
method robust only for halos with at least $10^3$ particles.  The
second problem is that there is not a way to estimate the optimal
radial bin size, different choices may produce different results for
the concentration.

The second method solves the two problems mentioned above.  
It works with low particle numbers and does not involve data binning.  
However, it effectively takes into account only a single data point
and discards the rest of the data.
Small fluctuations on the maximum can yield large perturbations on the
estimated concentration parameter.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{\emph{Left panel}. Average value of the fractional offset ($f_{\rm
    off}$ (Eq. \ref{eq:f_off}) between the concentration in a halo at 
  high resolution and its bootstrapped versions. 
  Thick (thin) lines correspond to massive halos in the Bolshoi (Via
  Lactea) simulation.
  The density method noticeably overestimates the concentration up to
  a factor of $20\%$, 
  while the new method only underestimates the concentrations by less
  than $3\%$.
  \emph{Right panel.}
  1$\sigma$ uncertainties on the offset and the concentration
  estimates from the integrated mass method.  
  Lines show the width between the 14 and 86 percentiles
  of the $f_{\rm off}$ distribution at fixed particle number. 
  The lines include the results for the three methods: maximum
  velocity, density and integrated mass.
  Grey symbols correspond to the $1\sigma$ uncertainty on each of
  the halos in the Bolshoi Simulation as a result of the MCMC
  sampling. 
  To allow a fair comparison with $f_{\rm off}$, this uncertainty has
  been normalized by the preferred concentration value for each halo. 
  \label{fig:downsampling}}
\end{figure*}

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.  The first method
takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  Then, it estimates the
density in each logarithmic bin.  At this point is possible to make a
direct fit to the density as a function of the radial coordinate.
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos. 
A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).

@@

Here we propose a new method to estimate the concentration. 
It uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).  We build it from N-body data as follows.
First, we define the center of the halo to be at the position of the
particle with the lowest gravitational potential. Then we rank the
particles by their increasing radial distance from the center.  From
this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. 
We then divide the enclosed mass $M_i$ and the radii $r_i$ by their
virial values to finally obtain the dimensionless variables $m_i$ and
$x_i$.

@@

If the integrated mass values at different radii were independent from
each other we could write a likelihood distribution as ${\cal
  L}(c|x_i)\propto \exp(-\chi^2(c,x_i)/2)$ with

\begin{equation}
\chi^2(c,x_i)= \sum_{i=2}^{N-1}\frac{[\log m_i - \log m(< x_i;c)]^2}{\sigma_i^2},
\end{equation}
%
where $\sigma_i^2=\sigma_x^2(x_i)$, $m(<x_i;c)$ corresponds to the
values in Eq.(\ref{eq:profile}) at $x=x_i$ for a given value of the
concentration parameter $c$, and the $i$ index sums over all the
particles in the numerical profile. 
In this computation the particles $i=1$ and $i=N$ are discarded to
avoid divergent terms in the sum.

However, tests on the bootstrapping data show that using
$\sigma_i^2=\sigma_x^2(x_i)$, instead of the full inverse covariance
matrix, grossly overestimates $\chi^2(c,x_i)$,
providing small uncertainties around the best concentration value. 
To avoid the expensive computation and inversion of a full covariance
matrix we use the bootsrapping data to calibrate an effective
$\sigma_i^2\approx \sigma_{\rm eff}^2(x_i)$.

We impose two conditions on the approximate $\sigma_{\rm eff}^{2}$. It
must keep the dependence on $x$ that we have discovered for the
diagonal elements and must give similar curves of $\chi^{2}(c,x_i)$-$c$
around the minimum as the full covariance matrix.
We found that this effective $\sigma_{\rm eff}^2$ can be approximated
as

@@

We then use an Affine Invariant Markov Chain Monte Carlo (MCMC)
implemented in the python module {\texttt{emcee}} \citep{emcee} to sample the
likelihood function distribution.
From the $\chi^2$ distribution we find the optimal value of the
concentration and its associated uncertainty. 
We stress that different choices for $\sigma^2_{\rm eff}$ do not
affect the optimal concentration value, only its uncertainty.

@@

We use two different simulations to test our methods.
The first is the Bolshoi run, a cosmological simulation that follows
the non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.   The
cosmological parameters use a Hubble parameter $h=0.73$, a matter
density $\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The data is publicly available at \url{http://www.cosmosim.org/}.  
Details about the structure of the database and the simulation can be
found in \citep{2011ApJ...740..102K,2013AN....334..691R}.

We use the halos located in a cubic sub-volume of $100$ \hMpc\ on a
side containing a total of $64531$ objects.
From this sample we select all the halos at $z=0$ detected with a
Friends-of-Friends (FoF) algorithm with more than 300 particles,
meaning that the masses are in the interval $4\times 10^{10}\leq
M_{\rm FoF}/\hMsun \leq 10^{14}$.  The FoF algorithm used a linking
length of $0.17$ times the mean inter-particle distance.  
This choice translates into an overdensity $\Delta_h\sim 400-700$
dependent on the halo concentration \citep{More2011}.

From this set of particles we follow the procedure spelled out in
Section \ref{sec:method} with $\Delta_h=740$ to select an spherical
region that we redefine to be our halo.  
This choice makes that the overdensities are fully included inside the
original FoF particle group.   
On the interest of providing a fair comparison against the density
method we only report results from overdensities with at least $200$
particles ($2.6\times 10^{10}$\hMsun).

We also use public data from the Via Lactea simulation project
\citep{2008Natur.454..735D}.  
This simulation contains a single isolated halo with a virial mass
of the order of $10^{12}$\hMsun simulated using the tree code PKDGRAV
\citep{2001PhDT........21S}.  
The simulation had $\sim 2\times 10^{8}$ particles to resolve this
region.
The cosmological parameters are different from those in the Bolshoi
simulation, with a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.238$ and a normalization of the power spectrum
$\sigma_8=0.74$. 
The data available to the public corresponds to a downsampled set 
of $10^5$ particles, which corresponds to a particle mass of
$2.24\times 10^{7}$\hMsun.

@@

We take halos with at least $10^{5}$ particles and subsample them by
factors of $2$ up to $10^{3}$. 
We measure the concentration at every resampling.
We use a two-sample Kolmogorov-Smirnov (KS) test to  compare the list
of radial distances from each subsample against that of its parent halo.
We find that the resulting p-value distribution is flat.
This confirms that the radial particle distribution in the bootraped
halo is consistent with coming from the distribution given by the
parent halo.

For every subsample we keep fixed the virial radius and the center
found for the high resolution halo.
Leaving the virial radius and center free in each bootstrapping
iteration has an effect smaller than $1\%$ in the concentration. 
In the Bolshoi simulation we select 14 massive halos and create 700
subsamples for each one.  
For the Via Lactea simulation the same halo is subsampled 10000
times.

@@

Figure \ref{fig:downsampling} summarizes our results.  
The plot on the left shows the average value of $f_{\rm off}$ as a
function of particle number.  
This can be interpreted as the statistical bias on the concentration
estimate. 
For large enough particle numbers, $N_{s}>4\times 10^3$ the results of
the three algorithms show a bias below the $1\%$ level.
For a lower number of particles the results start to deviate.
At $200$ particles the velocity method overestimates the concentration
by a factor $14\%$ while the density method overestimates it by $20\%$.
Around the same sampling scale, the new algorithm shows a more stable
behaviour underestimating the concentration only by a factor of
$1\%$-$3\%$.
Work  that studied the effect of
lowering numerical resolution on the same individual halos
\citep{2008MNRAS.391.1685S} yielded similar values for the
concentration across resolutions.  That work  explored a particle
number range $>5\times 10^5$, completely consistent with our result that the
bias can only be evident for particle numbers $<4\times 10^4$.

The thin lines on the same panel show a fit to the function 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}},
\label{eq:f_off_fit}
\end{equation} 
%
with $A=2842\pm 1900$, $B=7.96\pm 0.54$; $A=239\pm 131$, $B=6.23\pm
0.43$ and $A=-0.46\pm 3.49$, $B=0.79\pm 1.31$ for the density,
velocity and mass method, respectively.

The right panel in Figure \ref{fig:downsampling} shows different uncertainty
results.
The lines show the difference between the 14 and 86 percentiles in the
$f_{\rm off}$ distribution at fixed mass. 
Each line corresponds to the three different methods to estimate the
concentration and the two simulations.
This shows that the bootstrapping technique can help us to assign a
$1\sigma$ uncertainty to the concentration values at a fixed $N_s$as 
%
\begin{equation}
\sigma_{c} = \frac{0.40}{\sqrt{N_s/200}}.
\end{equation}
%

The circles in the same Figure show the $1\sigma$ uncertainty on all
the relaxed halos in the Bolshoi simulation sample using the MCMC
results.  
To allow for a fair comparison with the bootstrapping results, this
uncertainty is normalized to the concentration value.
The uncerainty from the bootstrapping experiment provides an upper
bound uncertainty on the concentration estimate for individual halos.

@@

We now inspect the mass-concentration relationship 
results with the three different algorithms.
This can help us to identify possible consequences of the 
biases detected through the bootstrapping experiments.

Figure \ref{fig:concentration} shows the mass-concentration
relationship for the density, velocity and integrated mass method.
The left panel shows the results as they are produced by each of the
algorithms. The thin dashed line marks the trend reported by
\citep{Prada2012} using the velocity method, showing that our
velocity method implementation can reproduce their results.

The results from the new algorithm follow very
closely the velocity algorithm at high masses ($M_h>10^{12}\hMsun$ or
equivalently for $>4\times10^3$ particles). 
For lower masses there is a difference between the median of the two
methods, but they are still consistent within the statistical
uncertainties.

@@

How do these results impact the most recent mass-concentration estimates? 
\cite{Ludlow2016} and \cite{Klypin2016} estimated the mass-concentration
relation over different suites of cosmological N-body simulations using the
density and velocity methods, respectively. 
Both used halos with at least $5\times 10^{3}$ particles.
This imposes a lower halo mass limit of $\sim 10^{12}$\hMsun (Figure 8
in \cite{Ludlow2016}, Figure 17 in \cite{Klypin2016}) to have robust
estimates. 
This means that their results for individual halos should not be
affected by the bias we report here.
However, this also leaves open the question about what other methods
can robustly say about the flattening we report below $10^{12}$\Msun
using the new method.

@@

In this letter we used bootstrapping to quantify the biases on
concentration estimates. 
We found that methods commonly used in the literature can overestimate
the concentrations by factors of $15\%$-$20\%$ for halos with $200$
particles, or $7\%$-$10\%$ for halos with $500$ particles.  
This procedure provides a robust technique to quantify the bias in
concentration estimates with the advantage that it works without
having to run new simulations.

These results motivated us to introduce a new method based on the
integrated mass profile that show a robust performance at low particle
numbers. 
The new algorithm showed a bias  of $< 3\%$ for halos with $200$
particles and less than $1\%$ for halos with $500$ particles or more.  
To keep the bias of the velocity and density methods below $2\%$ only halos
with at least $\sim 4000$ particles should be considered.

The three methods are in broad agreement, within the statistical
uncertainties, concerning their estimates of the  mass-concentration
relationship. 
Some noticeable differences include a $15\%$ systematically higher 
concentrations in the density method compared to the velocity method.
This systematic offset has been reported before with the same dataset 
\citep{Prada2012} and with different simulations \citep{Klypin2016}
without any conclusive explanation for its origin. 
Another difference is that the velocity and integrated mass methods
start to differ for masses below $10^{12}\hMsun$ ($\sim 4000$
particles).   
We found that correcting the mean concentration by the mean bias
factor found through bootstrapping brings these two techniques into
agreement.

These results show that using the integrated mass profile to estimate the
DM halo concentrations is a tool deserving deeper scrutiny.   
Further tests with larger simulated volumes, varying numerical
resolution, stacked data and different density profiles are the next
natural step to explore the full potential of this new method.

@@

 We acknowledge technical and financial support from Uniandes
 and Universidad de Antioquia.
 We acknowledge discussions with Tom\'as Verdugo, Stefan Gottloeber and Nelson
 Padilla.
 We thank the anonymous referee for comments that helped us improve
 the presentation of these results.

\bibliographystyle{apj}
%\bibliography{references}

\begin{thebibliography}{18}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

@@

\bibitem[{{Klypin} {et~al.}(2011){Klypin}, {Trujillo-Gomez}, \&
  {Primack}}]{2011ApJ...740..102K}
{Klypin}, A.~A., {Trujillo-Gomez}, S., \& {Primack}, J. 2011, \apj, 740, 102

\bibitem[{{Ludlow} {et~al.}(2016){Ludlow}, {Bose}, {Angulo}, {Wang},
  {Hellwing}, {Navarro}, {Cole}, \& {Frenk}}]{Ludlow2016}
{Ludlow}, A.~D., {Bose}, S., {Angulo}, R.~E., {Wang}, L., {Hellwing}, W.~A.,
  {Navarro}, J.~F., {Cole}, S., \& {Frenk}, C.~S. 2016, \mnras, 460, 1214

@@

\bibitem[{{Springel} {et~al.}(2008){Springel}, {Wang}, {Vogelsberger},
  {Ludlow}, {Jenkins}, {Helmi}, {Navarro}, {Frenk}, \&
  {White}}]{2008MNRAS.391.1685S}
{Springel}, V., {Wang}, J., {Vogelsberger}, M., {Ludlow}, A., {Jenkins}, A.,
  {Helmi}, A., {Navarro}, J.~F., {Frenk}, C.~S., \& {White}, S.~D.~M. 2008,
  \mnras, 391, 1685
